Gap formation and stability in non-isothermal protoplanetary discs

### 2014-11-21 paper/model.tex
penultimate edit by MKL

\section{Disc-planet models}\label{model}
The system is a two-dimensional non-self-gravitating gas disc orbiting
a central star of mass $M_*$.  
%self-gravitating gas disc of mass
%$M_d$ rotating about a central star of mass $M_*$. 
Cylindrical
co-ordinates $(r,\phi,z)$ centred on the star is adopted. The frame is   
non-rotating. Computational units are such that 
$G=M_*=\mathcal{R}=\mu=1$ where $G$ is the gravitational constant,
$\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight. %% \cite{lin12}

The disc is governed by the standard fluid equations  
\begin{align}\label{3d_gov_eq}
  &\frac{\p\Sigma}{\p t}+\nabla\cdot(\Sigma \bv)=0, \\
  & \frac{\p\bv}{\p t}+\bv\cdot\nabla\bv= -\frac{1}{\rho}\nabla p 
  - \nabla{\Phi} + \bm{f}_\nu,\\
  & \frac{\p e}{\p t} + \nabla\cdot(e\bv) = -p\nabla\cdot\bv +
  \mathcal{H} - \mathcal{C}, %\\
%  &\nabla^2\Phi_d = 4\pi G \Sigma \delta(z),\label{poisson}
\end{align}
where $\Sigma$ is the surface density, $\bv = (v_r,v_\phi)$ the fluid
velocity, $p$ is the vertically-integrated pressure, $e=p/(\gamma-1)$ is the energy
per unit area and the adiabatic index $\gamma=1.4$ is assumed
constant.

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) %, and the disc potential $\Phi_d$ which satisfies the
%Poisson equation \ref{poisson}. 
and indirect potentials to account for the non-inertial reference
frame.%  \citep[see][for explicit expressions]{masset02}
In the momentum 
equations, $\bm{f}_\nu$ represent viscous forces whose magnitude is 
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$ .

@@

%The surface density scale $\Sigma_0$ is set by  
%\begin{align}
%\Sigma_0 = \frac{h_i\rout\Omega_k^2(\rout)}{\pi G Q_o},
%\end{align}
%where $Q_o$ is the thin-disc Keplerian Toomre parameter at the outer boundary initially.

The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
pressure, stellar gravity and disc gravity. The initial radial velocity is $v_{r}=3\nu/r$ 
with $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$. For parameter values employed in this work, $v_{\phi}\simeq r\Omega_k$ 
and $\hat{\nu}= 10^{-9}$, so that $|v_{r}/v_{\phi}|\ll1$ and the initial 
flow is effectively only in the azimuthal direction.  With this choice
of physical viscosity $\hat{\nu}$ the only heating source is through
shock-heating (via artificial viscosity) and the $\mathcal{C}$
function when $e\Sigma<e_i\Sigma_i$.

@@

\begin{align}
  t_c(r_p+x_s) = \beta\Omega_k^{-1}(r_p+x_s) = \tilde{\beta} t_{\mathrm{lib}}(r_p+x_s), 
\end{align}
where $x_s=2r_h$ is the approximate distance from the planet to its
gap edge, and $t_\mathrm{lib}$ is the time interval between successive
encounters of a fluid element at the gap edge and the planet's
azimuth. That is, we measure the cooling time in units of the time
taken for a gap edge fluid element to encounter planet-induced
shocks.

Assuming Keplerian orbital frequencies and $x_s\ll r_p$
gives $t_\mathrm{lib}\simeq 4\pi r_p/(3\Omega_{kp} x_s)$, where
$\Omega_{kp} = \Omega_k(r_p)$. Therefore   
\begin{align}\label{betatilde}
  \beta = \tilde{\beta} \frac{4\pi r_p}{3x_s} \left(1  - \frac{3x_s}{2r_p}\right), 
\end{align}
where $x_s\ll r_p$ was used again. We use $x_s = 2r_h$ in
Eq. \ref{betatilde}. For a planet mass with $q=10^{-3}$,
Eq. \ref{betatilde} then gives $\beta \simeq 23.9\tilde{\beta}$.

%giant planet gap is 2.5rh
%2rh is CR of G.E.I. 
%\subsection{Simulation procedure}
%The disc is evolved without a planet for $t\in[0,20]P_0$. The planet potential is smoothly switched on
%between $t\in[20,30]P_0$.
%%  In a separate set of calculations the planet
%% potential is turned off after $t>t_\mathrm{off}$, immediately  
%% afterwhich the surface density, energy density and velocity fields are
%% azimuthally averaged and then subject to random surface density  
%% perturbations of magnitude $10^{-2}$.

@@

The generalized PV appears in the description of the linear stability
of radially-structured adiabatic discs \citep{lovelace99,li00}, where
the authors show an extremum in $\teta$ may lead to a dynamical
instability. In a barotropic disc, the entropy factor is absent
and the important quantity is the PV.

\subsubsection{Fourier modes} 
The Rossby wave instability is characterized by exponetially
growing pertibations. Though in this paper we do not formally consider
linear instability, modal analysis will be useful to
analyse the growth of different wavenumbers, which is associated with
the number of vortices initially formed by the RWI.

%characterize 
%azimuthal vortex numbers and overdensity values.

The Fourier transform of the time-dependent surface density is
\begin{align}\label{fouriertransform}
  \Sigma_m(r,t) = \int_{0}^{2\pi}
  \Sigma(r,\phi,t) \, \mathrm{e}^{-\mathrm{i}m\phi} \, \mathrm{d}\phi 
\end{align} 
where $m$ is the aziumuthal wave number. We define the growth rate
$\gamma$ of the $m^\mathrm{th}$ component of the surface density
through 
% and $\sigma_m$ is the complex frequency
%of the $m \mathrm{th}$ mode. Since $\sigma_m$ is a value dependent on
%$r$, a more useful diagnostic is the mode growth rate defined as  
\begin{align}\label{growth}
  \frac{d \langle|\Sigma_m|\rangle_r }{dt}= \gamma \langle|\Sigma_m|\rangle_r 
\end{align}
where %$\gamma$ is the real part of the $\sigma_m$ and 
$\langle|\cdot|\rangle_r$ denotes the average of the absolute value
over a raidal region of interest. By using Eq.~\ref{growth} the growth
rates of the unstable modes can be found from successive spatial
Fourier transforms over an appropriate period of time.

@@

%{\bf {MKL.} describe vortensity and generalized vortensity, and its
%  relevance to stability. fourier mode amplitude definition and how
%  growth rates are obtained. rossby number. density
%  perturbations. (see lin 2014)
%}

%% According to previous 2D studies, the key quantitiy in describing the
%% stability of radially structured non-barotropic discs is the
%% \emph{generalized vortensity} (GV): 
%% \begin{align}
%%   \tilde{\eta}\equiv \left(\frac{\kappa^2}{2\Omega\Sigma}\right)S^{-2/\gamma}, 
%% \end{align}
%% where $S \equiv p/\Sigma^\gamma$ is defined to be the entropy. The
%% quantity in brackets, $\eta\equiv \kappa^2/2\Omega\Sigma$, is the
%% usual vortensity or potential vorticity (PV).

>>>>

\section{Disc-planet models}\label{model}
The system is a two-dimensional (2D) non-self-gravitating gas disc orbiting
a central star of mass $M_*$. We adopt cylindrical
co-ordinates $(r,\phi,z)$ centred on the star. The frame is   
non-rotating. Computational units are such that 
$G=M_*=\mathcal{R}=\mu=1$ where $G$ is the gravitational constant,
$\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight.

The disc evolution is governed by the standard fluid equations  
\begin{align}\label{3d_gov_eq}
  &\frac{\p\Sigma}{\p t}+\nabla\cdot(\Sigma \bv)=0, \\
  & \frac{\p\bv}{\p t}+\bv\cdot\nabla\bv= -\frac{1}{\rho}\nabla p 
  - \nabla{\Phi} + \bm{f}_\nu,\\
  & \frac{\p e}{\p t} + \nabla\cdot(e\bv) = -p\nabla\cdot\bv +
  \mathcal{H} - \mathcal{C}, %\\
%  &\nabla^2\Phi_d = 4\pi G \Sigma \delta(z),\label{poisson}
\end{align}
where $\Sigma$ is the surface density, $\bv = (v_r,v_\phi)$ the fluid
velocity, $p$ is the vertically-integrated pressure, $e=p/(\gamma-1)$ is the energy
per unit area and the adiabatic index $\gamma=1.4$ is assumed
constant.

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum 
equations, $\bm{f}_\nu$ represent viscous forces whose magnitude is 
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$ .

@@

The initial azimuthal velocity $v_{\phi i}$ is set by centrifugal balance with
pressure, stellar gravity and disc gravity. For a thin disc, 
$v_{\phi}\simeq r\Omega_k$. The initial radial velocity is
$v_{r}=3\nu/r$, where $\nu = \hat{\nu}\rin^2\Omega_k(\rin)$, and we
adopt $\hat{\nu}= 10^{-9}$, so that $|v_{r}/v_{\phi}|\ll1$ and the initial 
flow is effectively only in the azimuthal direction.  With this value 
of physical viscosity, the only source of heating is through
compression, shock-heating (via artificial viscosity) and the
$\mathcal{C}$ function when $e/\Sigma<e_i/\Sigma_i$.

@@

\begin{align}
  t_c(r_p+x_s) = \beta\Omega_k^{-1}(r_p+x_s) = \tilde{\beta} t_{\mathrm{lib}}(r_p+x_s), 
\end{align}
where $x_s$ is the distance from the planet to its
gap edge, and $t_\mathrm{lib}$ is the time interval between successive
encounters of a fluid element at the gap edge and the planet's
azimuth. That is, we measure the cooling time in units of the time
interval between encounters of a fluid element at the gap edge and the
planet-induced shock.

Assuming Keplerian orbital frequencies and $x_s\ll r_p$
gives $t_\mathrm{lib}\simeq 4\pi r_p/(3\Omega_{kp} x_s)$, where
$\Omega_{kp} = \Omega_k(r_p)$. Therefore   
\begin{align}\label{betatilde}
  \beta = \tilde{\beta} \frac{4\pi r_p}{3x_s} \left(1  - \frac{3x_s}{2r_p}\right), 
\end{align}
where $x_s\ll r_p$ was used again. We use $x_s = 2r_h$ in
Eq. \ref{betatilde}. For a planet mass with $q=10^{-3}$,
Eq. \ref{betatilde} then gives $\beta \simeq 23.9\tilde{\beta}$. In
terms of planetary orbital periods, this is
\begin{align} 
  t_c(r) = \frac{\beta}{2\pi}\left(\frac{r}{r_p}\right)^{3/2}P_0\simeq
  3.8 \tilde{\beta}\left(\frac{r}{r_p}\right)^{3/2}P_0. 
\end{align}

@@

The generalized PV appears in the description of the linear stability
of radially-structured adiabatic discs \citep{lovelace99,li00}, where
the authors show an extremum in $\teta$ may lead to a dynamical
instability, the RWI. In a barotropic disc, the entropy factor is
absent and the important quantity is the PV.

\subsubsection{Fourier modes} 
The RWI is characterized by exponetially
growing perturbations. Though in this paper we do not consider a
formal linear instability calculation, modal analysis will be useful
to analyse the growth of perturbations with different azimuthal
wavenumbers, which is associated with the number of vortices initially
formed by the RWI.

The Fourier transform of the time-dependent surface density is
\begin{align}\label{fouriertransform}
  \Sigma_m(r,t) = \int_{0}^{2\pi}
  \Sigma(r,\phi,t) \, \mathrm{e}^{-\mathrm{i}m\phi} \, \mathrm{d}\phi 
\end{align} 
where $m$ is the aziumuthal wave number. We define the growth rate
$\gamma$ of the $m^\mathrm{th}$ component of the surface density
through 
% and $\sigma_m$ is the complex frequency
%of the $m \mathrm{th}$ mode. Since $\sigma_m$ is a value dependent on
%$r$, a more useful diagnostic is the mode growth rate defined as  
\begin{align}\label{growth}
  \frac{d \langle|\Sigma_m|\rangle_r }{dt}= \gamma \langle|\Sigma_m|\rangle_r 
\end{align}
where %$\gamma$ is the real part of the $\sigma_m$ and 
$\langle|\cdot|\rangle_r$ denotes the average of the absolute value
over a radial region of interest. By using Eq.~\ref{growth} the growth
rates of the unstable modes can be found from successive spatial
Fourier transforms over an appropriate period of time.

@@

### 2014-11-21 paper/paper.tex
penultimate edit by MKL

\maketitle
\begin{abstract}
%Hydrodynamic simulations of self-gravitating disc-planet interactions are presented. 
Several observations of transition discs show lopsided
dust-distributions. A potential explanation is the formation of a
large-scale vortex acting as a dust-trap at the edge of a gap opened
by a giant planet. Numerical models of gap-edge vortices have
thus far employed locally isothermal discs, but the 
theory of this vortex-forming or `Rossby wave' instability was
originally developed for adiabatic discs.  
We generalise the study of planetary gap stability to non-isothermal
discs using customised numerical simulations of disc-planet
systems where the planet open an unstable gap. 
We include in the energy equation a simple cooling term with
cooling timescale $t_\mathrm{cool}=\beta\Omega_k^{-1}$, where $\Omega_k$ is
the Keplerian frequency, and examine the effect of $\beta$ on the
stability of gap edges and vortex lifetimes. We find increasing
$\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
dominant azimuthal wavenumber $m$ decreases.  
%gap edges stabilize with increasing $\beta$ as dominate modes and
%growth rates decrease in value. 
We find a quasi-steady state consisting of one  
large-scale overdense vortex circulating the outer gap edge, typically  
lasting $O(10^3)$ orbits.  
%characterized by
%sudden disspation on the order of 100's or planet orbits.
Vortex lifetimes were found to increase with cooling times up to an
optimal value, after which vortex lifetimes decrease. This
non-monotonic dependence is qualitatively consistent with 
strictly isothermal simulations that differ in disc aspect ratios
presented in Fu et al..% (2014, ApJL, 788, L41).

%a non-monoticly dependence on $\beta$ and
%an optimum value for cooling rate was found to maximize the lifetime.

%The simulations are customized to examine the stability of gaps
%opened by massive planets as a function of $\beta$.

\end{abstract}

\begin{keywords}
planetary systems: formation --- planetary systems:
protoplanetary discs
\end{keywords}

@@

This paper is organised as follows. In \S\ref{model} we describe the
equations governing the disc-planet system and initial conditions. Our
numerical approach, including diagnositic measures, are given in
\S\ref{method}. We present results from two sets of numerical
experiments. In \S\ref{linear1} we use disc-planet interaction to set
up discs with gaps, but study their stability without further
influence from the planet. %stability through background state 
We then examine the lifetime of gap-edge vortices in \S\ref{nonlinear}
as a function of the imposed cooling rate. We conclude and summarize
in \S\ref{summary} with a discussion of important caveats.

>>>>

\maketitle
\begin{abstract}
  Several observations of transition discs show lopsided
  dust-distributions. A potential explanation is the formation of a
  large-scale vortex acting as a dust-trap at the edge of a gap opened
  by a giant planet. Numerical models of gap-edge vortices have
  thus far employed locally isothermal discs, but the 
  theory of this vortex-forming or `Rossby wave' instability was
  originally developed for adiabatic discs.  
  We generalise the study of planetary gap stability to non-isothermal
  discs using customised numerical simulations of disc-planet
  systems where the planet opens an unstable gap. 
  We include in the energy equation a simple cooling function with
  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
  the Keplerian frequency, and examine the effect of $\beta$ on the
  stability of gap edges and vortex lifetimes. We find increasing
  $\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
  dominant azimuthal wavenumber $m$ decreases.  
  We find a quasi-steady state consisting of one  
  large-scale, over-dense vortex circulating the outer gap edge, typically  
  lasting $O(10^3)$ orbits. 
  Vortex lifetimes were found to increase with cooling times up to an
  optimal value, beyond which vortex lifetimes decrease. This
  non-monotonic dependence is qualitatively consistent with 
  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
%  presented in Fu et al. (2014, ApJL, 788, L41). 
\end{abstract}

\begin{keywords}
  accretion, accretion discs, protoplanetary discs, hydrodynamics, instabilities,
  planet-disc interactions, methods: numerical 
\end{keywords}

@@

This paper is organised as follows. In \S\ref{model} we describe the
equations governing the disc-planet system and initial conditions. Our
numerical approach, including diagnositic measures, are given in
\S\ref{method}. We present results from two sets of numerical
experiments. In \S\ref{linear1} we use disc-planet interaction to set
up discs with gaps, but study their stability without further
influence from the planet. %stability through background state 
We then perform long-term disc-planet simulations to examine the
lifetime of gap-edge vortices in \S\ref{nonlinear},  
as a function of the imposed cooling rate. We conclude and summarize
in \S\ref{summary} with a discussion of important caveats.
